Parsing Postfix log files

### 2007-10-18 doc/logparser.tex
Changed the majority of \ref commands to my \refwithpage, which includes the page number automatically.  I like having the page number, but not how it interrupts the flow, so I think I'll have to work on the appearance.

@@

This section provides background information helpful in understanding the
remainder of the paper.  It begins with a discussion of the motivation
underlying the project, followed by some technical information: the use of
a database as an API\@; a brief introduction to the SMTP protocol; and a
longer introduction to Postfix, concentrating on the topics most relevant
to this paper, namely Postfix restrictions and policy servers.  The
assumptions made in designing and implementing the parser are explained, as
are the conventions used in this paper.  Other projects which attempt to
parse Postfix logs are summarised (full details are available in
Section~\ref{other-parsers}), plus a review of previously published
research in this area.

@@

This paper and the program it describes are part of a larger project to
optimise a server's Postfix restrictions, generate statistics and graphs,
and provide a platform on which new restrictions can be trialled and
evaluated to see if they are beneficial in the fight against spam.  The
program parses Postfix logs and populates a database with the data gleaned
from those logs, providing a consistent and simple view of the logs which
future tools can utilise.  The gathered data can then be used to optimise
current anti-spam measures, provide a baseline to test new anti-spam
measures against, or to produce statistics showing how effective those
measures are.\footnote{Section~\ref{introduction} lists some more example
uses.}

@@

Postfix's design follows the Unix philosophy of \textit{Write programs that
do one thing and do it well\/}~\cite{unix-philosophy}, and is separated
into various component programs to perform the tasks required of an MTA\@:
receive mail, send mail, local delivery of mail, etc. --- full details can
be found at~\cite{postfix-overview}.  Each log line contains the name of
the Postfix program which produced the log line, and this information is
used when determining which rules should be used to parse each log line
(see section~\ref{rule characteristics} for details.)

@@

10 other parsers have been reviewed in Appendix~\ref{other-parsers} as part
of the background research for this project.  None of the reviewed parsers
perform the type of advanced parsing and log correlation described here;
all are intended to perform a specific parsing and reporting task, rather
than be a generic parser, extracting data and leaving generation of reports
from the data to other programs.  Some parsers save the data extracted to a
database but the majority discard all data once they have finished running,
making historical analysis impossible.  The other parsers reviewed all
produce a report of greater or lesser complexity and detail, whereas the
program described here doesn't attempt to produce a report at all; that
responsibility is deferred to a separate program, to be developed later.
The parsing algorithm and program described here are designed to enable
much more detailed log analysis by providing a stable platform for
subsequent programs to develop upon.

@@

Parsing in LMA is hand-coded and requires major changes to the code to
parse new log lines or extract additional data.  It does not appear to deal
with any of the complications discussed in
sections~\ref{complications}~and~\ref{additional complications}.  The
program fails to parse Postfix 2.2.x or 2.3.x log files; indeed it produces
a number of warnings and no usable output.  Parsing in \parsername{} is
divided in two (see sections~\ref{rules} and~\ref{parsing-algorithm} for a
more detailed explanation): parsing a new log line is a trivial matter of
adding a new rule, but code changes may be required if new complications
(sections~\ref{complications} and~\ref{additional complications}) must be
dealt with or a new action is required (section~\ref{actions-in-detail}).

@@

The database schema can be conceptually divided in two: the rules which are
used to parse log files, and the data saved from the parsing of log files.
Rules have the fields required to parse the log lines, extract data to be
saved, and the action to be executed; they also have several fields which
aid the user in understanding what each rule parses.  The rules are
described in detail in section~\ref{rules} but the fields are covered in
section~\ref{rule attributes}.

The data saved from parsing the logs is also divided into two tables as
described below: connections and results.  The connections table contains a
row for every mail accepted and every connection where there was a
rejection; the individual fields will be described in the forthcoming
section~\ref{connections table}.  The results table has one or more rows
per row in the connections table, depending on the specifics of the
mail/connection; the fields will be covered in detail in
section~\ref{results table}.

@@

Rules are discussed in detain in section~\ref{rules}, but the rules table
is covered here.  Rules are created by the user, not the parser, and will
not be modified by the parser (except for the hits and hits\_total fields).
Rules parse the individual lines, telling the parser which fields to
extract and what action to take for that line.

@@

        For efficiency the keywords are expanded and every rule's regex is
        compiled before attempting to parse the log file --- otherwise each
        regex would be recompiled each time it was used, resulting in a
        large, data dependent slowdown.  Rule efficiency concerns are
        discussed in section~\ref{rule efficiency}.

@@

    \item [action] The action the algorithm will take; a full list can be
        found in Section~\ref{actions-in-detail}.

@@

    \item [hits] is an efficiency measure.  This counter is maintained for
        every rule and incremented each time the rule successfully matches.
        At the start of each run the program sorts the rules in descending
        order of hits, and at the end of the run updates every rule's hits.
        Assuming that the distribution of log lines is reasonably
        consistent between log files, rules matching more commonly
        occurring log lines will be tried before rules matching less
        commonly occurring log lines, lowering the program's execution
        time.  Rule ordering for efficiency is discussed in
        section~\ref{rule ordering for efficiency}.

@@

    \item [priority] This is the user-configurable companion to hits: rules
        will be tried in order of priority, overriding hits.  This allows
        more specific rules to take precedence over more general rules
        (described in section~\ref{overlapping rules}).

@@

This section discusses the rules used in parsing Postfix log files,
starting with rule characteristics, continuing with detecting and dealing
with overlapping rules, and the problems they can cause.  An example rule
and a line it would match are provided, plus a description of how the
fields in the rule are used when matching and performing the action (the
details of the table containing the rules have already been described in
section~\ref{rule attributes},).  This section finishes with a discussion
of rule efficiency concerns, referring to the graphs in
section~\ref{graphs}.

@@

    \item Compare the results of parsing using sorted, shuffled and
        reversed rules.\footnote{See section~\ref{rule efficiency} for more
        details of sorting the rules.}  Parse a number of log files using
        normal sorting, dump a textual representation of the connections
        and results tables, and delete everything from those tables.
        Repeat with shuffled and reversed sorting.  If there are no
        overlapping rules the tables from each run will be identical;
        differences indicate overlapping rules, and the rule\_ids which
        differ determine which rules overlap.  Unfortunately this method
        cannot prove the absence of overlapping rules; it can detect
        overlapping rules --- but only if there are log lines in the input
        files which match more than one rule.

@@

    \item [action:] will be executed if the regex matches successfully (see
        section~\ref{actions-in-detail} for full details).

@@

    \item [hits, hits\_total and priority:] hits and priority are used in
        ordering the rules (see section~\ref{rule ordering for
        efficiency}); hits\_total is updated at the end of each run but is
        otherwise unused.

@@

Parsing efficiency is an obvious concern when the parser routinely needs to
deal with 75 MB log files containing 300,000 log lines (for a small mail
server --- large scale mail servers would have much larger log files on a
daily basis).  When generating the data for the graphs included in
section~\ref{graphs}, 93 log files (totaling 10.08 GB, 60.72 million lines)
were each parsed 10 times, the first run discarded and the remaining 9 runs
averaged.  The first run is discarded for two reasons:

@@

An important property of a parser is how execution time scales with input
size: does it scale linearly, polynomially, or exponentially?
Graph~\ref{execution time vs file size vs number of lines graph} shows the
execution time in seconds, file size in MB and tens of thousands of lines
per log file per log file.  All three lines run roughly in parallel, giving
a visual impression that the algorithm scales linearly with input size.
This impression is borne out by graph~\ref{execution time vs file size vs
number lines factor}, which plots the ratio of file size vs execution time
and number of lines vs execution time.  As the reader can see, the ratios
are quite tightly banded, showing that the algorithm scales linearly: the
much larger log files between points 60 and 70 on the X axis in
graph~\ref{execution time vs file size vs number of lines graph} don't
cause any abnormality in the corresponding points in graph~\ref{execution
time vs file size vs number lines factor}.

@@

Rule ordering was mentioned in section~\ref{rule attributes} and will be
covered in greater detail in this section.  At the time of writing there
are 139 different rules, with the top 10\% matching the vast majority of
the log lines, and the remaining log lines split across the other 90\% of
the rules (as shown in graph~\ref{rule hits}).  Assuming that the
distribution of log lines is reasonably steady over time, program
efficiency should benefit from trying more frequently matching rules before
those which match less frequently.  To test this hypothesis three full test
runs were performed with different rule orderings:

@@

Graphs~\ref{percentage increase of shuffled over normal}
and~\ref{percentage increase of reversed over normal} show the percentage
increase of execution times (mean and standard deviation are shown after
the graph).  Overall this provides a modest but worthwhile performance
increase.

@@

Perl compiles the original regex into an internal representation,
optimising the regex to improve the speed of matching, but this compilation
and optimisation takes CPU time.  Perl automatically caches static regexs,
but dynamic regexs need to be explicitly compiled and cached.
Graph~\ref{normal regex vs discard regex} shows execution times with and
without caching the regex.  Caching the compiled regexs is obviously far
more efficient; graph~\ref{normal regex vs discarded regex factor} shows
the percentage execution time increase when not caching each regex.

@@

    \item The database table containing the rules is dealt with in
        section~\ref{rule attributes}, and is not duplicated in this
        section.

    \item The final part of this section discusses rule efficiency,
        covering the effects of caching compiled regexs and optimal
        ordering ordering of rules, with reference to the graphs in
        section~\ref{graphs}.

@@

Multiple independent mails may be delivered during one connection: this
requires cloning the current data as soon as a mail is accepted, so that
subsequent mails won't trample over each other's data.  This must be done
every time a mail is accepted, as it's impossible to tell in advance which
connections will accept multiple mails.  It is quite easy to overlook this
complication because only a small minority of connections accept more than
one mail. Happily once the mail has been accepted log entries won't be
correlated by pid for that mail any more (its queueid will be used
instead), so there isn't any ambiguity about which mail a given log line
belongs to.\footnote{Unfortunately this statement is not completely
accurate: see section~\ref{timeouts-during-data-phase} for details.
However in general there isn't any ambiguity about which data structure
should be used for a given log line.}  The original connection will be
discarded unsaved when the client disconnects if it doesn't have any data
worth saving, i.e.\ no rejections.  One unsolved difficulty is
distinguishing between different groups of rejections, e.g.\ when dealing
with the following sequence:

@@

Process scheduling can have a still more confusing effect: quite often the
child mail will be created, delivered and entirely finished with
\textbf{before} the parent logs the creation line!  Thus mails flagged as
coming from an unknown origin cannot be entered into the database when
their final log line is parsed; instead they must be marked as ready for
entry and subsequently entered by the parent mail once it has been
identified.  Quite apart from identifying mail injected in an unknown
fashion, or bugs in the parser, unknown origin mails need to be marked as
such because they lack some data present in the parent mail; they must copy
that data from their parent before being entered in the database.  The
effect of this complication upon the algorithm is discussed in
section~\ref{tracking re-injected mail}.

@@

Section~\ref{actions-in-detail} explains the actions in substantive detail;
this section will omit such detail because it would clutter and confuse the
algorithm description.  The data flow chart in section~\ref{flow-chart}
should also be consulted while reading this section.  For the sake of
clarity this description covers the most common paths through the
algorithm, as including every path would hinder understanding.  The more
uncommon paths are caused by the complications described in
section~\ref{additional complications}, and are covered therein.

@@

    \item If one or more mails were accepted there will be more log entries
        for those mails later, see section~\ref{mail-delivery}.

@@

Re-injection due to forwarding sadly lacks explicit log lines of its
own;\footnote{Previously discussed in section~\ref{complications},
complication 3.} re-injection is somewhat awkward to explain because it
overlaps both the mail acceptance and mail delivery sections, so discussion
is deferred to section~\ref{tracking re-injected mail}.

Internally generated mails lack any explicit origin in Postfix 2.2.x and
must be detected using heuristics (see
section~\ref{identifying-bounce-notifications} for details).  Bounce
notifications are the primary example of internally generated mails, though
there may be other types.

@@

    \item Sometimes mail is re-injected and the child mail needs to be
        tracked by the parent mail (action: track; flowchart:~11) ---
        section~\ref{tracking re-injected mail} discusses this in detail.

    \item Eventually the mail will be delivered, bounced, or deleted by the
        administrator (action: commit; flowchart:~12).  This is the last
        log line for this particular mail (though it may be indirectly
        referred to if it was re-injected).  If it is neither parent nor
        child of re-injection the data is cleaned up and entered in the
        database (flowchart:~14), then deleted from the state tables
        (flowchart:~15).  Re-injected mails are described in
        section~\ref{tracking re-injected mail}.

@@

It should be reiterated that the actions above happen whether the mail is
delivered to a mailbox, piped to a command, delivered to a remote server,
bounced (due to a mail loop, delivery failure, or five day timeout), or
deleted by the administrator.  The exception is what happens after delivery
to the parent or children of mail re-injected due to forwarding, and that
is explained in section~\ref{tracking re-injected mail}.

@@

    \item The indicator which more commonly introduces re-injection is when 
        \texttt{postfix/qmgr}
        selects a mail with a previously unseen queueid for delivery
        (action: mail\_picked\_for\_delivery; flowchart:~3), in which case
        a new data structure will be created.  The mail will be flagged as
        having unknown origins; this flag should be subsequently cleared
        once the origin has been established.  This may also be an
        indicator that the mail is a bounce notification, see
        section~\ref{identifying-bounce-notifications} for details.

@@

    \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
        picking a mail from the queue to deliver. This action is used for
        both qmgr and cleanup as it needs to deal with out of order log
        lines; see section~\ref{discarding cleanup lines} for details.

@@

        Out of order log entries may have caused the state table entry to
        already exist (see section~\ref{pickup logging after cleanup});
        otherwise it is created.  The data from the log line is then saved
        to the state table entry.

@@

    \item [TIMEOUT] The connection timed out so the mail currently being
        transferred must be discarded. The mail may have been accepted, in
        which case there's a data structure to dispose of, or it may not in
        which case there is not.  See
        section~\ref{timeouts-during-data-phase} for the gory details.

@@

Postfix 2.2.x (and presumably previous versions) lacks explicit logging
when bounce notifications are generated; suddenly there will be log entries
for a mail which lacks an obvious source.  There are similarities to the
problem of re-injected mails discussed in
section~\ref{tracking re-injected mail}, but unlike the solution described
therein bounce notifications do not eventually have a log line which
identifies their source.  Heuristics must be used to identify bounce
notifications, and those heuristics are:

@@

If both checks are successful then the mail is assumed to be one of the
offending bogus mails and is discarded.  There will be no further entries
logged for such mails, so without identifying and discarding them they
accumulate in the state table and will cause clashes if the queueid is
reused.  The mail cannot be entered in the database as the only data
available is the client hostname and IP address, but the database schema
requires many more fields be populated (see section~\ref{database schema}).

@@

Some mail clients disconnect abruptly if a second or subsequent recipient
is rejected; they may also disconnect after other errors, but such
disconnections are either unimportant or are handled elsewhere in the
algorithm (section~\ref{timeouts-during-data-phase}).  Sadly Postfix
doesn't log a message saying the mail has been discarded, as should be
expected by now.  The checks to identify this happening are:

@@

This complication is further complicated by out of order cleanup lines: see
section~\ref{discarding cleanup lines} for details.

@@

The above appear to cover all situations where an smtpd suddenly stops
logging.  In addition to removing an active connection the last accepted
mail may need to be discarded, as detailed in
section~\ref{timeouts-during-data-phase}.

@@

Occasionally a log file will have out of order log lines which cannot be
dealt with by the techniques described in sections~\ref{tracking
re-injected mail},~\ref{discarding cleanup lines}~and~\ref{pickup logging
after cleanup}.  In the 93 log files used for testing this occurs only five
times in 60,721,709 log lines, but for completeness of the algorithm it
should be dealt with.  The five occurrences in the test log files have the
same characteristics: the \texttt{postfix/local} log line showing delivery
to a local mailbox occurs after the \texttt{postfix/qmgr} log line showing
removal of the mail from the queue because delivery is completed.  This
causes problems: the mail is not complete, so entry into the database
fails; a new mail is created when the \texttt{postfix/local} line is parsed
and remains in the state tables; four warnings are issued per pair of out
of order log lines.

@@

The aborted delivery attempts described in
section~\ref{aborted-delivery-attempts} occur frequently, but the aborted
delivery attempts described in this section only occur four times in the 93
log files used for testing.  The symptoms are the same as in
section~\ref{aborted-delivery-attempts}, except that there \textit{is\/} a
\texttt{postfix/cleanup} log line; there does not appear to be anything in
the log file to explain why there are no further log messages.  The only
way to detect these mails is to periodically scan all mails in the state
tables, deleting any mails displaying the following characteristics:

@@

    \item The appropriate action to take --- this is discussed in
        section~\ref{mails-covered}.

@@

Full coverage of log lines is easy to achieve yet hard to maintain.  It is
easy to achieve full coverage for a limited set of log files (at the time
of writing the parser has 139 rules, fully parsing 93 contiguous log files
from Postfix 2.2 and 2.3), and new rules are easy to add.  Maintaining full
coverage is hard because other servers have different restrictions with
custom messages, RBLs change their messages over time, major releases of
Postfix change warning messages (usually adding more information), etc.\ so
over time the log lines drift and change.  Graph~\ref{rule hits} shows the
number of hits for each rule over all 93 log files; it's obvious that a
small number of rules match the vast majority of the lines, and more than
half the rules match fewer than 100 times.

@@

    \item It appears that the hostname used in the HELO command is not
        logged if the mail is accepted.\footnote{Tested with Postfix 2.2.10
        and 2.3.11; this may possibly have changed in Postfix 2.4.}
        Rectifying this has already been described in section~\ref{logging
        helo}.

@@

        Lire does attempt to correlate log lines by queueid, not pid, but
        its hard coded parsing extracts only part of the available data and
        makes no attempt to deal with the other complications described in
        sections~\ref{complications} and~\ref{additional complications}.
        It stores data, but not in an easily accessible format.  It
        generates reports in multiple formats, but those reports do not
        appear to be customisable.

>>>>

% Replacement for \ref{}, adds the page number too.
\newcommand{\refwithpage}[1]{%
    \empty{}\ref{#1} (page~\pageref{#1})%
}

@@

This section provides background information helpful in understanding the
remainder of the paper.  It begins with a discussion of the motivation
underlying the project, followed by some technical information: the use of
a database as an API\@; a brief introduction to the SMTP protocol; and a
longer introduction to Postfix, concentrating on the topics most relevant
to this paper, namely Postfix restrictions and policy servers.  The
assumptions made in designing and implementing the parser are explained, as
are the conventions used in this paper.  Other projects which attempt to
parse Postfix logs are summarised (full details are available in
section~\refwithpage{other-parsers}), plus a review of previously published
research in this area.

@@

This paper and the program it describes are part of a larger project to
optimise a server's Postfix restrictions, generate statistics and graphs,
and provide a platform on which new restrictions can be trialled and
evaluated to see if they are beneficial in the fight against spam.  The
program parses Postfix logs and populates a database with the data gleaned
from those logs, providing a consistent and simple view of the logs which
future tools can utilise.  The gathered data can then be used to optimise
current anti-spam measures, provide a baseline to test new anti-spam
measures against, or to produce statistics showing how effective those
measures are.\footnote{Section~\refwithpage{introduction} lists some more
example uses.}

@@

Postfix's design follows the Unix philosophy of \textit{Write programs that
do one thing and do it well\/}~\cite{unix-philosophy}, and is separated
into various component programs to perform the tasks required of an MTA\@:
receive mail, send mail, local delivery of mail, etc. --- full details can
be found at~\cite{postfix-overview}.  Each log line contains the name of
the Postfix program which produced the log line, and this information is
used when determining which rules should be used to parse each log line
(see section~\refwithpage{rule characteristics} for details.)

@@

10 other parsers have been reviewed in Appendix~\refwithpage{other-parsers}
as part of the background research for this project.  None of the reviewed
parsers perform the type of advanced parsing and log correlation described
here; all are intended to perform a specific parsing and reporting task,
rather than be a generic parser, extracting data and leaving generation of
reports from the data to other programs.  Some parsers save the data
extracted to a database but the majority discard all data once they have
finished running, making historical analysis impossible.  The other parsers
reviewed all produce a report of greater or lesser complexity and detail,
whereas the program described here doesn't attempt to produce a report at
all; that responsibility is deferred to a separate program, to be developed
later.  The parsing algorithm and program described here are designed to
enable much more detailed log analysis by providing a stable platform for
subsequent programs to develop upon.

@@

Parsing in LMA is hand-coded and requires major changes to the code to
parse new log lines or extract additional data.  It does not appear to deal
with any of the complications discussed in
sections~\refwithpage{complications} and~\refwithpage{additional
complications}.  The program fails to parse Postfix 2.2.x or 2.3.x log
files; indeed it produces a number of warnings and no usable output.
Parsing in \parsername{} is divided in two (see
sections~\refwithpage{rules} and~\refwithpage{parsing-algorithm} for a more
detailed explanation): parsing a new log line is a trivial matter of adding
a new rule, but code changes may be required if new complications
(sections~\refwithpage{complications} and~\refwithpage{additional
complications}) must be dealt with or a new action is required
(section~\refwithpage{actions-in-detail}).

@@

The database schema can be conceptually divided in two: the rules which are
used to parse log files, and the data saved from the parsing of log files.
Rules have the fields required to parse the log lines, extract data to be
saved, and the action to be executed; they also have several fields which
aid the user in understanding what each rule parses.  The rules are
described in detail in section~\refwithpage{rules} but the fields are
covered in section~\refwithpage{rule attributes}.

The data saved from parsing the logs is also divided into two tables as
described below: connections and results.  The connections table contains a
row for every mail accepted and every connection where there was a
rejection; the individual fields will be described in the forthcoming
section~\refwithpage{connections table}.  The results table has one or more
rows per row in the connections table, depending on the specifics of the
mail/connection; the fields will be covered in detail in
section~\refwithpage{results table}.

@@

Rules are discussed in detain in section~\refwithpage{rules}, but the rules
table is covered here.  Rules are created by the user, not the parser, and
will not be modified by the parser (except for the hits and hits\_total
fields).  Rules parse the individual lines, telling the parser which fields
to extract and what action to take for that line.

@@

        For efficiency the keywords are expanded and every rule's regex is
        compiled before attempting to parse the log file --- otherwise each
        regex would be recompiled each time it was used, resulting in a
        large, data dependent slowdown.  Rule efficiency concerns are
        discussed in section~\refwithpage{rule efficiency}.

@@

    \item [action] The action the algorithm will take; a full list can be
        found in Section~\refwithpage{actions-in-detail}.

@@

    \item [hits] is an efficiency measure.  This counter is maintained for
        every rule and incremented each time the rule successfully matches.
        At the start of each run the program sorts the rules in descending
        order of hits, and at the end of the run updates every rule's hits.
        Assuming that the distribution of log lines is reasonably
        consistent between log files, rules matching more commonly
        occurring log lines will be tried before rules matching less
        commonly occurring log lines, lowering the program's execution
        time.  Rule ordering for efficiency is discussed in
        section~\refwithpage{rule ordering for efficiency}.

@@

    \item [priority] This is the user-configurable companion to hits: rules
        will be tried in order of priority, overriding hits.  This allows
        more specific rules to take precedence over more general rules
        (described in section~\refwithpage{overlapping rules}).

@@

This section discusses the rules used in parsing Postfix log files,
starting with rule characteristics, continuing with detecting and dealing
with overlapping rules, and the problems they can cause.  An example rule
and a line it would match are provided, plus a description of how the
fields in the rule are used when matching and performing the action (the
details of the table containing the rules have already been described in
section~\refwithpage{rule attributes},).  This section finishes with a
discussion of rule efficiency concerns, referring to the graphs in
section~\refwithpage{graphs}.

@@

    \item Compare the results of parsing using sorted, shuffled and
        reversed rules.\footnote{See section~\refwithpage{rule efficiency}
        for more details of sorting the rules.}  Parse a number of log
        files using normal sorting, dump a textual representation of the
        connections and results tables, and delete everything from those
        tables.  Repeat with shuffled and reversed sorting.  If there are
        no overlapping rules the tables from each run will be identical;
        differences indicate overlapping rules, and the rule\_ids which
        differ determine which rules overlap.  Unfortunately this method
        cannot prove the absence of overlapping rules; it can detect
        overlapping rules --- but only if there are log lines in the input
        files which match more than one rule.

@@

    \item [action:] will be executed if the regex matches successfully (see
        section~\refwithpage{actions-in-detail} for full details).

@@

    \item [hits, hits\_total and priority:] hits and priority are used in
        ordering the rules (see section~\refwithpage{rule ordering for
        efficiency}); hits\_total is updated at the end of each run but is
        otherwise unused.

@@

Parsing efficiency is an obvious concern when the parser routinely needs to
deal with 75 MB log files containing 300,000 log lines (for a small mail
server --- large scale mail servers would have much larger log files on a
daily basis).  When generating the data for the graphs included in
section~\refwithpage{graphs}, 93 log files (totaling 10.08 GB, 60.72
million lines) were each parsed 10 times, the first run discarded and the
remaining 9 runs averaged.  The first run is discarded for two reasons:

@@

An important property of a parser is how execution time scales with input
size: does it scale linearly, polynomially, or exponentially?
Graph~\refwithpage{execution time vs file size vs number of lines graph}
shows the execution time in seconds, file size in MB and tens of thousands
of lines per log file per log file.  All three lines run roughly in
parallel, giving a visual impression that the algorithm scales linearly
with input size.  This impression is borne out by
graph~\refwithpage{execution time vs file size vs number lines factor},
which plots the ratio of file size vs execution time and number of lines vs
execution time.  As the reader can see, the ratios are quite tightly
banded, showing that the algorithm scales linearly: the much larger log
files between points 60 and 70 on the X axis in
graph~\refwithpage{execution time vs file size vs number of lines graph}
don't cause any abnormality in the corresponding points in
graph~\refwithpage{execution time vs file size vs number lines factor}.

@@

Rule ordering was mentioned in section~\refwithpage{rule attributes} and
will be covered in greater detail in this section.  At the time of writing
there are 139 different rules, with the top 10\% matching the vast majority
of the log lines, and the remaining log lines split across the other 90\%
of the rules (as shown in graph~\refwithpage{rule hits}).  Assuming that
the distribution of log lines is reasonably steady over time, program
efficiency should benefit from trying more frequently matching rules before
those which match less frequently.  To test this hypothesis three full test
runs were performed with different rule orderings:

@@

Graphs~\refwithpage{percentage increase of shuffled over normal}
and~\refwithpage{percentage increase of reversed over normal} show the
percentage increase of execution times (mean and standard deviation are
shown after the graph).  Overall this provides a modest but worthwhile
performance increase.

@@

Perl compiles the original regex into an internal representation,
optimising the regex to improve the speed of matching, but this compilation
and optimisation takes CPU time.  Perl automatically caches static regexs,
but dynamic regexs need to be explicitly compiled and cached.
Graph~\refwithpage{normal regex vs discard regex} shows execution times
with and without caching the regex.  Caching the compiled regexs is
obviously far more efficient; graph~\refwithpage{normal regex vs discarded
regex factor} shows the percentage execution time increase when not caching
each regex.

@@

    \item The database table containing the rules is dealt with in
        section~\refwithpage{rule attributes}, and is not duplicated in
        this section.

    \item The final part of this section discusses rule efficiency,
        covering the effects of caching compiled regexs and optimal
        ordering ordering of rules, with reference to the graphs in
        section~\refwithpage{graphs}.

@@

Multiple independent mails may be delivered during one connection: this
requires cloning the current data as soon as a mail is accepted, so that
subsequent mails won't trample over each other's data.  This must be done
every time a mail is accepted, as it's impossible to tell in advance which
connections will accept multiple mails.  It is quite easy to overlook this
complication because only a small minority of connections accept more than
one mail. Happily once the mail has been accepted log entries won't be
correlated by pid for that mail any more (its queueid will be used
instead), so there isn't any ambiguity about which mail a given log line
belongs to.\footnote{Unfortunately this statement is not completely
accurate: see section~\refwithpage{timeouts-during-data-phase} for details.
However in general there isn't any ambiguity about which data structure
should be used for a given log line.}  The original connection will be
discarded unsaved when the client disconnects if it doesn't have any data
worth saving, i.e.\ no rejections.  One unsolved difficulty is
distinguishing between different groups of rejections, e.g.\ when dealing
with the following sequence:

@@

Process scheduling can have a still more confusing effect: quite often the
child mail will be created, delivered and entirely finished with
\textbf{before} the parent logs the creation line!  Thus mails flagged as
coming from an unknown origin cannot be entered into the database when
their final log line is parsed; instead they must be marked as ready for
entry and subsequently entered by the parent mail once it has been
identified.  Quite apart from identifying mail injected in an unknown
fashion, or bugs in the parser, unknown origin mails need to be marked as
such because they lack some data present in the parent mail; they must copy
that data from their parent before being entered in the database.  The
effect of this complication upon the algorithm is discussed in
section~\refwithpage{tracking re-injected mail}.

@@

Section~\refwithpage{actions-in-detail} explains the actions in substantive
detail; this section will omit such detail because it would clutter and
confuse the algorithm description.  The data flow chart in
section~\refwithpage{flow-chart} should also be consulted while reading
this section.  For the sake of clarity this description covers the most
common paths through the algorithm, as including every path would hinder
understanding.  The more uncommon paths are caused by the complications
described in section~\refwithpage{additional complications}, and are
covered therein.

@@

    \item If one or more mails were accepted there will be more log entries
        for those mails later, see section~\refwithpage{mail-delivery}.

@@

Re-injection due to forwarding sadly lacks explicit log lines of its
own;\footnote{Previously discussed in section~\refwithpage{complications},
complication 3.} re-injection is somewhat awkward to explain because it
overlaps both the mail acceptance and mail delivery sections, so discussion
is deferred to section~\refwithpage{tracking re-injected mail}.

Internally generated mails lack any explicit origin in Postfix 2.2.x and
must be detected using heuristics (see
section~\refwithpage{identifying-bounce-notifications} for details).
Bounce notifications are the primary example of internally generated mails,
though there may be other types.

@@

    \item Sometimes mail is re-injected and the child mail needs to be
        tracked by the parent mail (action: track; flowchart:~11) ---
        section~\refwithpage{tracking re-injected mail} discusses this in
        detail.

    \item Eventually the mail will be delivered, bounced, or deleted by the
        administrator (action: commit; flowchart:~12).  This is the last
        log line for this particular mail (though it may be indirectly
        referred to if it was re-injected).  If it is neither parent nor
        child of re-injection the data is cleaned up and entered in the
        database (flowchart:~14), then deleted from the state tables
        (flowchart:~15).  Re-injected mails are described in
        section~\refwithpage{tracking re-injected mail}.

@@

It should be reiterated that the actions above happen whether the mail is
delivered to a mailbox, piped to a command, delivered to a remote server,
bounced (due to a mail loop, delivery failure, or five day timeout), or
deleted by the administrator.  The exception is what happens after delivery
to the parent or children of mail re-injected due to forwarding, and that
is explained in section~\refwithpage{tracking re-injected mail}.

@@

    \item The indicator which more commonly introduces re-injection is when
        \texttt{postfix/qmgr} selects a mail with a previously unseen
        queueid for delivery (action: mail\_picked\_for\_delivery;
        flowchart:~3), in which case a new data structure will be created.
        The mail will be flagged as having unknown origins; this flag
        should be subsequently cleared once the origin has been
        established.  This may also be an indicator that the mail is a
        bounce notification, see
        section~\refwithpage{identifying-bounce-notifications} for details.

@@

    \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
        picking a mail from the queue to deliver. This action is used for
        both qmgr and cleanup as it needs to deal with out of order log
        lines; see section~\refwithpage{discarding cleanup lines} for
        details.

@@

        Out of order log entries may have caused the state table entry to
        already exist (see section~\refwithpage{pickup logging after
        cleanup}); otherwise it is created.  The data from the log line is
        then saved to the state table entry.

@@

    \item [TIMEOUT] The connection timed out so the mail currently being
        transferred must be discarded. The mail may have been accepted, in
        which case there's a data structure to dispose of, or it may not in
        which case there is not.  See
        section~\refwithpage{timeouts-during-data-phase} for the gory
        details.

@@

Postfix 2.2.x (and presumably previous versions) lacks explicit logging
when bounce notifications are generated; suddenly there will be log entries
for a mail which lacks an obvious source.  There are similarities to the
problem of re-injected mails discussed in section~\refwithpage{tracking
re-injected mail}, but unlike the solution described therein bounce
notifications do not eventually have a log line which identifies their
source.  Heuristics must be used to identify bounce notifications, and
those heuristics are:

@@

If both checks are successful then the mail is assumed to be one of the
offending bogus mails and is discarded.  There will be no further entries
logged for such mails, so without identifying and discarding them they
accumulate in the state table and will cause clashes if the queueid is
reused.  The mail cannot be entered in the database as the only data
available is the client hostname and IP address, but the database schema
requires many more fields be populated (see section~\refwithpage{database
schema}).

@@

Some mail clients disconnect abruptly if a second or subsequent recipient
is rejected; they may also disconnect after other errors, but such
disconnections are either unimportant or are handled elsewhere in the
algorithm (section~\refwithpage{timeouts-during-data-phase}).  Sadly
Postfix doesn't log a message saying the mail has been discarded, as should
be expected by now.  The checks to identify this happening are:

@@

This complication is further complicated by out of order cleanup lines: see
section~\refwithpage{discarding cleanup lines} for details.

@@

The above appear to cover all situations where an smtpd suddenly stops
logging.  In addition to removing an active connection the last accepted
mail may need to be discarded, as detailed in
section~\refwithpage{timeouts-during-data-phase}.

@@

Occasionally a log file will have out of order log lines which cannot be
dealt with by the techniques described in sections~\refwithpage{tracking
re-injected mail},~\refwithpage{discarding cleanup
lines}~and~\refwithpage{pickup logging after cleanup}.  In the 93 log files
used for testing this occurs only five times in 60,721,709 log lines, but
for completeness of the algorithm it should be dealt with.  The five
occurrences in the test log files have the same characteristics: the
\texttt{postfix/local} log line showing delivery to a local mailbox occurs
after the \texttt{postfix/qmgr} log line showing removal of the mail from
the queue because delivery is completed.  This causes problems: the mail is
not complete, so entry into the database fails; a new mail is created when
the \texttt{postfix/local} line is parsed and remains in the state tables;
four warnings are issued per pair of out of order log lines.

@@

The aborted delivery attempts described in
section~\refwithpage{aborted-delivery-attempts} occur frequently, but the
aborted delivery attempts described in this section only occur four times
in the 93 log files used for testing.  The symptoms are the same as in
section~\refwithpage{aborted-delivery-attempts}, except that there
\textit{is\/} a \texttt{postfix/cleanup} log line; there does not appear to
be anything in the log file to explain why there are no further log
messages.  The only way to detect these mails is to periodically scan all
mails in the state tables, deleting any mails displaying the following
characteristics:

@@

    \item The appropriate action to take --- this is discussed in
        section~\refwithpage{mails-covered}.

@@

Full coverage of log lines is easy to achieve yet hard to maintain.  It is
easy to achieve full coverage for a limited set of log files (at the time
of writing the parser has 139 rules, fully parsing 93 contiguous log files
from Postfix 2.2 and 2.3), and new rules are easy to add.  Maintaining full
coverage is hard because other servers have different restrictions with
custom messages, RBLs change their messages over time, major releases of
Postfix change warning messages (usually adding more information), etc.\ so
over time the log lines drift and change.  Graph~\refwithpage{rule hits}
shows the number of hits for each rule over all 93 log files; it's obvious
that a small number of rules match the vast majority of the lines, and more
than half the rules match fewer than 100 times.

@@

    \item It appears that the hostname used in the HELO command is not
        logged if the mail is accepted.\footnote{Tested with Postfix 2.2.10
        and 2.3.11; this may possibly have changed in Postfix 2.4.}
        Rectifying this has already been described in
        section~\refwithpage{logging helo}.

@@

        Lire does attempt to correlate log lines by queueid, not pid, but
        its hard coded parsing extracts only part of the available data and
        makes no attempt to deal with the other complications described in
        sections~\refwithpage{complications} and~\refwithpage{additional
        complications}.  It stores data, but not in an easily accessible
        format.  It generates reports in multiple formats, but those
        reports do not appear to be customisable.
